PyUltraLight: A Pseudo-Spectral Solver for Ultralight Dark Matter Dynamics

### 2018-05-15 draft.tex
changes from Richard; tweaks to format to work well with standalone LaTeX, tweaks to refs, and rewrite of abstract/intro

\usepackage[justification=justified]{caption}
\usepackage{varwidth}
\usepackage{graphicx}

\title{\boldmath PyUltraLight: A Pseudo-Spectral Code for the Dynamics of Ultra-Light Dark Matter}

@@

\abstract{P\textsc{y}U\textsc{ltra}L\textsc{ight} is a Python based program that solves the Schr{\"o}dinger-Poisson coupled differential equations, which govern the dynamics of ultra-light scalar field dark matter in the non-relativistic regime. P\textsc{y}U\textsc{ltra}L\textsc{ight} utilises a symmetrised split-step pseudo-spectral Fourier algorithm to solve these differential equations, and demonstrates excellent energy conservation. Being Python-based, the program is particularly easy to set up, and the accompanying Jupyter notebook provides a simple interface in which to augment simulation parameters and create visualisations of output data. Using P\textsc{y}U\textsc{ltra}L\textsc{ight}, we are able to reproduce many previously studied dynamical characteristics of ultra-light dark matter, even when running the simulations at relatively low spatial resolution from a standard user-PC.}

@@

The true nature of dark matter (DM) is one of the most important outstanding problems in theoretical physics. WIMP models of dark matter are favoured by many, due in part to their consistency with supersymmetric extensions to the standard model of particle physics. However, with increasingly advanced direct-detection experiments failing to find evidence for WIMP candidates, the viable parameter space continues to shrink \cite{Tan2016, Akerib2017}. Furthermore, the $\Lambda$CDM cosmologies consistent with WIMP models suffer from a number of widely-publicised small-scale problems \cite{Bull2016}. While baryonic physics is proposed to offer solutions to a number of these problems, the perceived issues within $\Lambda$CDM are increasingly motivating novel research into alternative DM models, one such class of model being ultra-light dark matter (ULDM), also known as fuzzy dark matter (FDM), or BEC dark matter. ULDM models are well motivated by fundamental theories possessing approximate shift symmetries, and offer natural resolutions to the small-scale problems of $\Lambda$CDM without the need to invoke baryonic physics. This is essentially because below the de Broglie wavelength of the ULDM particle, the Heisenberg uncertainty principle suppresses gravitational collapse. Since the mass of the ULDM particle is postulated to be of order $10^-22$ eV, this amounts to suppression of gravitational collapse on kilo parsec scales \cite{Hu2000}. A recent review of ULDM/FDM is given in \cite{Hui2016}.

@@

While these modified cosmological simulation codes have produced promising results in the study of ULDM physics, at the time of writing only one such code, \textsc{axion-gadget}, has been made publicly available. The aim of this paper is to introduce P\textsc{y}U\textbf{ltra}L\textsc{ight}, a stand-alone Python-based pseudo-spectral SP solver, and to demonstrate that this code is able to reproduce a number of key findings from the more complicated cosmological simulation codes without the need for high-performance computing environments. Consequently, we anticipate that as a publicly available resource, P\textsc{y}U\textbf{ltra}L\textsc{ight} will serve as a valuable cross-check for more complex implementations and may serve as a basis for further development of such codes within the computational cosmology community.

P\textsc{y}U\textbf{ltra}L\textsc{ight} makes use of a symmetrised-split-step pseudo-spectral Fourier algorithm (also known as leap-frog) to compute the dynamics of the SP system. We note that a similar methodology has been described in \cite{Paredes2016}, though to the authors' knowledge this code has not been released publicly. The symmetrised-split-step method offers $2^{nd}$ order accurate time integration steps and sub-percent level energy conservation while normalisation of the wavefunction is conserved to machine precision. As a pseudo-spectral code, linear differential operators are computed by direct multiplication in the Fourier domain, while non-linear terms are evaluated in position space. This means that P\textsc{y}U\textbf{ltra}L\textsc{ight} is free from noise associated with finite-differencing methods of computing spatial derivatives, though there is a necessary computational cost associated with the Fourier and inverse Fourier transforms. These transforms are optimised in P\textsc{y}U\textbf{ltra}L\textsc{ight} through the use of the pyFFTW pythonic wrapper around the C-based FFWTW subroutine library \cite{pyfftw} \cite{fftw}. While the code is currently implemented to run on a singe machine, the FFTW libraries offer full parallelisation, so there is no barrier to running the code in a cluster environment.

This paper is organised as follows. We first provide a short review of the physics of ULDM, including a derivation of the SP equations from an appropriate scalar-field Lagrangian. We then describe the details of the implementation of P\textsc{y}U\textbf{ltra}L\textsc{ight}, before moving on to a testing and verification section, in which we reproduce a selection of results from a variety of recent ULDM simulations and discuss energy conservation within the code.

@@

The existence of an extremely light scalar field, minimally coupled to gravity, is the central premise on which ULDM models are predicated. Within the ULDM framework, it is proposed that large-scale cosmological structure formation proceeds through the Bose-Einstein condensation of this scalar field. In this section we derive the (non-relativistic) dynamical equations of the single macroscopic condensate wavefunction, namely, the Schr{\"o}dinger-Poisson equations. We begin by considering the following action functional for a scalar field, $\phi$, minimally coupled to gravity and in the absence of self-interactions:
\begin{equation}\label{eq:action}
    S=\int \frac{d^4x}{\hbar}\sqrt{-g}\bigg\{\frac{1}{2}g^{\mu\nu}\partial_\mu\phi\partial_\nu\phi-\frac{m^2}{2\hbar^2}\phi^2\bigg\},
\end{equation}
where we have taken $c=1$ but retain factors of $\hbar$ at this stage. Applying the variational principle to this action yields the Euler-Lagrange equations
\begin{equation}\label{eq:e-l}
    \frac{1}{\sqrt{-g}}\partial_\mu\big[\sqrt{-g}g^{\mu\nu}\partial_\nu\phi\big]-\frac{m^2}{\hbar^2}\phi=0.
\end{equation}
We evaluate equation \ref{eq:e-l} using linear perturbation theory, adopting the perturbed FRW metric in the Newtonian gauge:
\begin{equation}\label{eq:pFRW}
    ds^2=-\big(1+2\Phi(\vec{r})\big)dt^2+a(t)^2\big(1-2\Phi(\vec{r})\big)d\vec{r}^{\,2}.
\end{equation}
To linear order in $\Phi(\vec{r})$ we obtain 
\begin{equation}\label{eq:linear}
    \Ddot{\phi}-\frac{\big(1+4\Phi(\vec{r})\big)}{a(t)^2}\nabla^2\phi+3\big(1+2\Phi(\vec{r})\big)H\Dot{\phi}+\big(1+2\Phi(\vec{r})\big)\frac{m^2}{\hbar^2}\phi=0,
\end{equation}
where $H=\Dot{a}(t)/a(t)$. At late times, $H\ll m$ and it is sufficient to set $H=0$ and $a(t)=1$ in equation \ref{eq:linear}. In the non-relativistic limit, we may use the WKB approximation to obtain an ansatz solution to equation \ref{eq:linear}. We take 
\begin{equation}\label{eq:ansatz}
    \phi=\frac{\hbar}{\sqrt{2}m}\big(\psi e^{-imt/\hbar}+\psi^* e^{imt/\hbar}\big),
\end{equation}
where $\psi$ is a slowly varying complex function so that terms involving $\Ddot{\psi}$ may be neglected. We may also use $\nabla^2\phi\ll m^2/\hbar^2\phi$, such that our final result takes the form
\begin{equation}\label{eq:schrodinger}
    i\hbar\Dot{\psi}=-\frac{\hbar^2}{2m}\nabla^2\psi+m\Phi\psi.
\end{equation}
Thus, by making suitable approximations we have derived the Schr{\"o}dinger equation for $\psi$, which we now interpret as the unique macroscopic wavefunction of a Bose-Einstein condensate. It follows that the particle density of the condensate is given by $\vert\psi\vert^2$, such that the mass density is simply $m\vert\psi\vert^2$. Using this, we may therefore write down the Poisson equation as
\begin{equation}\label{eq:poisson}
    \nabla^2\Phi=4\pi G m \vert\psi\vert^2,
\end{equation}
where $G$ is Newton's gravitational constant. The coupled equations \ref{eq:schrodinger} and \ref{eq:poisson} together form the nonlinear Schr{\"o}dinger-Poisson (SP) system which describes the dynamics of ULDM in the non-relativistic regime. It is this system of equations which P\textsc{y}U\textsc{ltra}L\textsc{ight} is designed to solve.

@@

It is helpful to recast the SP system (equations \ref{eq:schrodinger} and \ref{eq:poisson}) in terms of adimensional quantities. In keeping with \cite{Schive2014} and \cite{Paredes2016}, we introduce length, time, and mass scales as follows:
\begin{align}
    \CMcal{L}&=\left(\frac{8\pi\hbar^2}{3 m^2H_0^2\Omega_{m_0}}\right)^{\frac{1}{4}}\approx121\left(\frac{10^{-23}\operatorname{eV}}{m}\right)\operatorname{kpc},\label{eq:length}\\
    \CMcal{T}&=\left(\frac{8\pi}{3 H_0^2\Omega_{m_0}}\right)^{\frac{1}{2}}\approx75.5 \operatorname{Gyr},\label{eq:time}\\
    \CMcal{M}&=\frac{1}{G}\left(\frac{8\pi}{3 H_0^2\Omega_{m_0}}\right)^{-\frac{1}{4}}\left(\frac{\hbar}{m}\right)^{\frac{3}{2}}\approx 7\times 10^7\left(\frac{10^{-23}\operatorname{eV}}{m}\right)\operatorname{M}_{\odot},\label{eq:mass}
\end{align}
where $m$ is the mass of the ultra-light scalar field, $H_0$ is the present-day Hubble parameter, $G$ is Newton's gravitational constant and $\Omega_{m_0}$ is the present-day matter fraction of the energy density of the universe. We now recast equations \ref{eq:schrodinger} and \ref{eq:poisson} in terms of the following dimensionless quantities:
\begin{equation}\label{eq:dimensionless}
    t'=\frac{t}{\CMcal{T}},\quad
    \vec{x}^{\,'}=\frac{\vec{x}}{\CMcal{L}},\quad
    \Phi'=\frac{m\CMcal{T}}{\hbar}\Phi,\quad
    \psi'=\CMcal{T}\sqrt{mG}\psi.
\end{equation}
Dropping the primes for notational convenience, we see that the coupled differential equations of the SP system are reduced to 
\begin{align}
    i\Dot{\psi}(\vec{x},t)&=-\frac{1}{2}\nabla^2\psi(\vec{x},t)+\Phi(\vec{x},t)\psi(\vec{x},t),\label{eq:s-adim}\\
    \nabla^2\Phi(\vec{x},t)&=4\pi\vert\psi(\vec{x},t)\vert^2,\label{eq:p-adim}
\end{align}
where it is understood that all quantities involved are dimensionless. We can recover any dimensionful quantities of interest through simple transformations using equations \ref{eq:length} to \ref{eq:mass}. For example, the integrated mass of the system, $M_{tot}$, is given by
\begin{equation}\label{eq:integrated-mass}
    M_{tot}=\CMcal{M}\int d^3x\vert\psi\vert^2.
\end{equation}
Likewise, the mass density at any point is given by
\begin{equation}\label{eq:density}
    \rho=\frac{\CMcal{M}}{\CMcal{L}^3}\vert\psi\vert^2.
\end{equation}
All computations performed by P\textsc{y}U\textsc{ltra}L\textsc{ight} are carried out using the dimensionless quantities defined in \ref{eq:dimensionless}, while provisions are made within the code for initial conditions and outputs to be converted to and from the desired unit system specified by the user. Henceforth, we will often refer to the quantity $\vert\psi\vert^2$ simply as the density, where it is understood that this is in fact a dimensionless quantity related to the physical mass density through a constant of proportionality given by equation \ref{eq:density}.

@@

Dynamical evolution within P\textsc{y}U\textsc{ltra}L\textsc{ight} progresses via a symmetrised split-step Fourier process on an $N\times N\times N$ simulation grid with spatial periodic boundary conditions. To understand this method, we first consider the exact expression for the unitary time evolution of the wavefunction according to equation \ref{eq:s-adim}, namely
\begin{equation}\label{eq:exact-time-ev}
    \psi(\vec{x},t+h)=\mathcal{T}\operatorname{exp}\left[-i\int_t^{t+h}dt'\left\{-\frac{1}{2}\nabla^2+\Phi(\vec{x},t')\right\}\right]\psi(\vec{x},t),
\end{equation}
where $\mathcal{T}$ is the time-ordering symbol. For sufficiently small timestep $h$, we can use the trapezoidal rule make the approximation
\begin{equation}
    \int_t^{t+h}dt'\Phi(\vec{x},t')\approx \frac{h}{2}\Big(\Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)+\Phi(\vec{x},t)\Big).
\end{equation}
We can therefore write the approximate form of equation \ref{eq:exact-time-ev} as
\begin{equation}\label{eq:approx-time-ev}
    \psi(\vec{x},t+h)\approx \operatorname{exp}\left[i\frac{h}{2}\Big(\nabla^2-\Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)-\Phi(\vec{x},t)\Big)\right]\psi(\vec{x},t).
\end{equation}
It is acknowledged that the exponential in equation \ref{eq:approx-time-ev} omits the time-ordering symbol, and is only equivalent to its time-ordered counterpart to order $h^2$.

@@

Evaluation of equation \ref{eq:split-step} within P\textsc{y}U\textsc{ltra}L\textsc{ight} proceeds as follows:
Initially, the nonlinear term acts in position space for one half-timestep. The result is then Fourier transformed, and a full timestep is taken in which the differential operator is applied in the Fourier domain. The potential field is then updated in accordance with equation \ref{eq:p-adim}. An inverse Fourier transform is then performed, and a final half timestep is taken in which the updated nonlinear term acts in position space to give the new $\psi$ field configuration. This procedure is known as the symmetrised split-step Fourier method, and is used widely in fields such as nonlinear fiber optics \cite{Agrawal2013}. It can be represented schematically as follows:
\begin{equation}\label{eq:schematic-psi}
    \psi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)=\operatorname{exp}\left[-\frac{ih}{2}\Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)\right]\CMcal{F}^{-1} \operatorname{exp}\left[\frac{-ih}{2}k^2\right]\CMcal{F}\operatorname{exp}\left[-\frac{ih}{2}\Phi(\vec{x},t)\right]\psi(\vec{x},t),
\end{equation}
where the order of operations is read from right to left, $\CMcal{F}$ and $\CMcal{F}^{-1}$ represent the discrete Fourier transform and its inverse, and $k$ is the wavenumber in the Fourier domain. The potential field is updated following the inverse Fourier transform in equation \ref{eq:schematic-psi}, and this update procedure may be represented schematically as
\begin{equation}\label{eq:schematic-phi}
    \Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)=\CMcal{F}^{-1}\left(-\frac{1}{k^2}\right)\CMcal{F}\ 4\pi\vert\psi(\vec{x},t_{i})\vert^2,
\end{equation}
where $\psi(\vec{x},t_{i})$ represents the field configuration at this intermediary point in the timestep. The final operation in equation \ref{eq:schematic-psi} changes only the phase of the $\psi$ field, and thus we could replace $\psi(\vec{x},t_{i})$ with $\psi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)$ in equation \ref{eq:schematic-phi} with no change in meaning.

In the P\textsc{y}U\textsc{ltra}L\textsc{ight} implementation, we make use of an additional simplification of the symmetrised split-step Fourier method. Namely, consecutive half-steps in the nonlinear term may be combined into a single full step in each case such that when iterated, only the first and last operations involve half steps. This can be represented schematically as 
\begin{equation}
    \psi(t\hspace*{-.2em}+\hspace*{-.2em}nh)=\operatorname{exp}\left[+\frac{ih}{2}\Phi\right]\left(\prod^n\operatorname{exp}\left[-\frac{ih}{2}\Phi\right]\operatorname{exp}\left[\frac{ih}{2}\nabla^2\right]\right)\operatorname{exp}\left[-\frac{ih}{2}\Phi\right]\psi(t),
\end{equation}
where the $\Phi$ field is updated at each step in accordance with equation \ref{eq:schematic-phi} and attention is drawn to the sign difference between the first and last operators.

The discrete Fourier transform (DFT) and its inverse are handled within P\textsc{y}U\textsc{ltra}L\textsc{ight} using pyFFTW. pyFFTW is a pythonic wrapper around the C-based FFTW subroutine library, and allows for efficient evaluation of both real and complex DFTs \cite{pyfftw} \cite{fftw} \cite{Frigo2005}.  Multithreading is supported by pyFFTW, and is implemented within P\textsc{y}U\textsc{ltra}L\textsc{ight} by using the multiprocessing package to determine the number of online CPUs and feed this directly to the threads argument of the pyfftw.FFTW class.

In addition to the use of pyFFTW, performance within P\textsc{y}U\textsc{ltra}L\textsc{ight} is further improved through the use of the NumExpr package, which provides routines for the efficient manipulation of array objects within the simulation using a vector-based virtual machine \cite{numexpr}.

@@

The initialisation procedure within P\textsc{y}U\textsc{ltra}L\textsc{ight} can be easily augmented to accommodate the requirements of the user. In the current implementation, the initial field configuration is built by loading a NumPy array file encoding a solitonic solution to the Schr{\"o}dinger-Poisson system. Using this method, solitons may be placed at the desired locations within the simulation grid, with the corresponding mass, velocity, and phase parameters each specified by the user within the accompanying Jupyter notebook.

@@

Soliton collisions can be classified into two categories depending on whether the total energy of the isolated binary system is positive or negative. In the case of positive energy, the solitons pass through each other, emerging largely undisturbed from their initial configurations following the interaction. During the collision, the wavefunctions describing the solitons are superposed, yielding distinctive interference patterns. In this section we demonstrate the ability of P\textsc{y}U\textsc{ltra}L\textsc{ight} to reproduce the interference patterns predicted by theoretical models during binary collisions.

>>>>

\usepackage[justification=justified]{caption}

\title{ PyUltraLight: A Pseudo-Spectral Solver for Ultralight Dark Matter Dynamics}

@@

\newcommand{\apj}{Astrophysical J.}
\newcommand{\prd}{Phys. Rev. D}
\newcommand{\mnras}{Mon. Notices Royal Astron. Soc}
\newcommand{\PyUltraLight}{\textsc{PyUltraLight}}

\abstract{\PyUltraLight\ simulates the dynamics of ultralight dark matter in a static, non-expanding background.  \PyUltraLight\  can describe the evolution of several interacting ultralight dark matter halos  or one or more halos orbiting a central, fixed Newtonian potential, the latter scenario corresponding to dwarf galaxies orbiting a massive central galaxy. We verify  \PyUltraLight\  by showing that it reproduces qualitative dynamical features of previously published simulations and demonstrate that it has  excellent energy-conservation properties.   \PyUltraLight\  is implemented in a Python-based Jupyter notebook, solving the Schr{\"o}dinger-Poisson equation  governing  ultralight scalar field dark matter dynamics in the non-relativistic regime using a symmetrised split-step pseudo-spectral  algorithm. The notebook interface makes it simple to specify simulation parameters and  visualise the resulting output. \PyUltraLight\  runs on standard desktop hardware and is available on GitHub.  }

@@

The realisation that there may be more to the universe than meets the eye is one of the most profound developments in 20$^{\textrm{th}}$ Century astronomy and astrophysics. However,  while there are now multiple lines of evidence that dark matter outweighs baryonic matter by a ratio of approximately 5:1 [CITE], we have few clues regarding the physical nature of dark matter.  Much theoretical and experimental effort has focussed on  WIMP models, motivated by their consistency with supersymmetric extensions to the Standard Model and their relatively simple dynamics but advanced direct-detection experiments are putting increasingly tight constraints on the WIMP parameter space \cite{Tan2016, Akerib2017}. Moreover, $\Lambda$CDM cosmology with simple, pressureless, noninteracting dark matter (a class which includes simple WIMP models) is potentially at odds with observations at small astrophysical scales \cite{Bull2016}.

[TO HERE]The potential shortcoming of simple cold dark matter scenarios are increasingly motivating the investigations of more novel dark matter scenarios. In particular, ultralight dark matter (ULDM) is an increasingly well-studied possibility, also known as fuzzy dark matter (FDM), or BEC dark matter. ULDM models are well motivated by fundamental theories possessing approximate shift symmetries, and offer natural resolutions to the small-scale problems of $\Lambda$CDM without the need to invoke baryonic physics. This is essentially because below the de Broglie wavelength of the ULDM particle, the Heisenberg uncertainty principle suppresses gravitational collapse. Since the mass of the ULDM particle is postulated to be of order $10^-22$ eV, this amounts to suppression of gravitational collapse on kilo parsec scales \cite{Hu2000}. A recent review of ULDM/FDM is given in \cite{Hui2016}.

@@

While these modified cosmological simulation codes have produced promising results in the study of ULDM physics, at the time of writing only one such code, \textsc{axion-gadget}, has been made publicly available. The aim of this paper is to introduce \PyUltraLight, a stand-alone Python-based pseudo-spectral SP solver, and to demonstrate that this code is able to reproduce a number of key findings from the more complicated cosmological simulation codes without the need for high-performance computing environments. Consequently, we anticipate that as a publicly available resource, \PyUltraLight\ will serve as a valuable cross-check for more complex implementations and may serve as a basis for further development of such codes within the computational cosmology community.

\PyUltraLight\ makes use of a symmetrised-split-step pseudo-spectral Fourier algorithm (also known as leap-frog) to compute the dynamics of the SP system. We note that a similar methodology has been described in \cite{Paredes2016}, though to the authors' knowledge this code has not been released publicly. The symmetrised-split-step method offers $2^{nd}$ order accurate time integration steps and sub-percent level energy conservation while normalisation of the wavefunction is conserved to machine precision. As a pseudo-spectral code, linear differential operators are computed by direct multiplication in the Fourier domain, while non-linear terms are evaluated in position space. This means that \PyUltraLight\ is free from noise associated with finite-differencing methods of computing spatial derivatives, though there is a necessary computational cost associated with the Fourier and inverse Fourier transforms. These transforms are optimised in \PyUltraLight\ through the use of the pyFFTW pythonic wrapper around the C-based FFWTW subroutine library \cite{pyfftw} \cite{fftw}. While the code is currently implemented to run on a singe machine, the FFTW libraries offer full parallelisation, so there is no barrier to running the code in a cluster environment.

This paper is organised as follows. We first provide a short review of the physics of ULDM, including a derivation of the SP equations from an appropriate scalar-field Lagrangian. We then describe the details of the implementation of \PyUltraLight, before moving on to a testing and verification section, in which we reproduce a selection of results from a variety of recent ULDM simulations and discuss energy conservation within the code.

@@

The existence of an extremely light scalar field, minimally coupled to gravity, is the central premise on which ULDM models are predicated. Within the ULDM framework, it is proposed that large-scale cosmological structure formation proceeds through the Bose-Einstein condensation of this scalar field. In this section we derive the (non-relativistic) dynamical equations of the single macroscopic condensate wavefunction, namely, the Schr{\"o}dinger-Poisson equations. We begin by considering the following action functional for a scalar field, $\phi$, minimally coupled to gravity and in the absence of self-interactions:
\begin{equation}\label{eq:action}
    S=\int \frac{d^4x}{\hbar}\sqrt{-g}\bigg\{\frac{1}{2}g^{\mu\nu}\partial_\mu\phi\partial_\nu\phi-\frac{m^2}{2\hbar^2}\phi^2\bigg\},
\end{equation}
%
where we have taken $c=1$ but retain factors of $\hbar$ at this stage. Applying the variational principle to this action yields the Euler-Lagrange equations
%
\begin{equation}\label{eq:e-l}
    \frac{1}{\sqrt{-g}}\partial_\mu\big[\sqrt{-g}g^{\mu\nu}\partial_\nu\phi\big]-\frac{m^2}{\hbar^2}\phi=0.
\end{equation}
We evaluate equation \ref{eq:e-l} using linear perturbation theory, adopting the perturbed FRW metric in the Newtonian gauge:
\begin{equation}\label{eq:pFRW}
    ds^2=-\big(1+2\Phi(\vec{r})\big)dt^2+a(t)^2\big(1-2\Phi(\vec{r})\big)d\vec{r}^{\,2}.
\end{equation}
To linear order in $\Phi(\vec{r})$ we obtain 
\begin{equation}\label{eq:linear}
    \Ddot{\phi}-\frac{\big(1+4\Phi(\vec{r})\big)}{a(t)^2}\nabla^2\phi+3\big(1+2\Phi(\vec{r})\big)H\Dot{\phi}+\big(1+2\Phi(\vec{r})\big)\frac{m^2}{\hbar^2}\phi=0,
\end{equation}
where $H=\Dot{a}(t)/a(t)$. At late times, $H\ll m$ and it is sufficient to set $H=0$ and $a(t)=1$ in equation \ref{eq:linear}. In the non-relativistic limit, we may use the WKB approximation to obtain an ansatz solution to equation \ref{eq:linear}. We take 
\begin{equation}\label{eq:ansatz}
    \phi=\frac{\hbar}{\sqrt{2}m}\big(\psi e^{-imt/\hbar}+\psi^* e^{imt/\hbar}\big),
\end{equation}
where $\psi$ is a slowly varying complex function so that terms involving $\Ddot{\psi}$ may be neglected. We may also use $\nabla^2\phi\ll m^2/\hbar^2\phi$, such that our final result takes the form
\begin{equation}\label{eq:schrodinger}
    i\hbar\Dot{\psi}=-\frac{\hbar^2}{2m}\nabla^2\psi+m\Phi\psi.
\end{equation}
Thus, by making suitable approximations we have derived the Schr{\"o}dinger equation for $\psi$, which we now interpret as the unique macroscopic wavefunction of a Bose-Einstein condensate. It follows that the particle density of the condensate is given by $\vert\psi\vert^2$, such that the mass density is simply $m\vert\psi\vert^2$. Using this, we may therefore write down the Poisson equation as
\begin{equation}\label{eq:poisson}
    \nabla^2\Phi=4\pi G m \vert\psi\vert^2,
\end{equation}
where $G$ is Newton's gravitational constant. The coupled equations \ref{eq:schrodinger} and \ref{eq:poisson} together form the nonlinear Schr{\"o}dinger-Poisson (SP) system which describes the dynamics of ULDM in the non-relativistic regime. It is this system of equations which \PyUltraLight is designed to solve.

@@

It is helpful to recast the SP system (equations \ref{eq:schrodinger} and \ref{eq:poisson}) in terms of adimensional quantities. In keeping with \cite{Schive2014} and \cite{Paredes2016}, we introduce length, time, and mass scales as follows:
\begin{align}
    \CMcal{L}&=\left(\frac{8\pi\hbar^2}{3 m^2H_0^2\Omega_{m_0}}\right)^{\frac{1}{4}}\approx121\left(\frac{10^{-23}\operatorname{eV}}{m}\right)\operatorname{kpc},\label{eq:length}\\
    \CMcal{T}&=\left(\frac{8\pi}{3 H_0^2\Omega_{m_0}}\right)^{\frac{1}{2}}\approx75.5 \operatorname{Gyr},\label{eq:time}\\
    \CMcal{M}&=\frac{1}{G}\left(\frac{8\pi}{3 H_0^2\Omega_{m_0}}\right)^{-\frac{1}{4}}\left(\frac{\hbar}{m}\right)^{\frac{3}{2}}\approx 7\times 10^7\left(\frac{10^{-23}\operatorname{eV}}{m}\right)\operatorname{M}_{\odot},\label{eq:mass}
\end{align}
where $m$ is the mass of the ultralight scalar field, $H_0$ is the present-day Hubble parameter, $G$ is Newton's gravitational constant and $\Omega_{m_0}$ is the present-day matter fraction of the energy density of the universe. We now recast equations \ref{eq:schrodinger} and \ref{eq:poisson} in terms of the following dimensionless quantities:
\begin{equation}\label{eq:dimensionless}
    t'=\frac{t}{\CMcal{T}},\quad
    \vec{x}^{\,'}=\frac{\vec{x}}{\CMcal{L}},\quad
    \Phi'=\frac{m\CMcal{T}}{\hbar}\Phi,\quad
    \psi'=\CMcal{T}\sqrt{mG}\psi.
\end{equation}
Dropping the primes for notational convenience, we see that the coupled differential equations of the SP system are reduced to 
\begin{align}
    i\Dot{\psi}(\vec{x},t)&=-\frac{1}{2}\nabla^2\psi(\vec{x},t)+\Phi(\vec{x},t)\psi(\vec{x},t),\label{eq:s-adim}\\
    \nabla^2\Phi(\vec{x},t)&=4\pi\vert\psi(\vec{x},t)\vert^2,\label{eq:p-adim}
\end{align}
where it is understood that all quantities involved are dimensionless. We can recover any dimensionful quantities of interest through simple transformations using equations \ref{eq:length} to \ref{eq:mass}. For example, the integrated mass of the system, $M_{tot}$, is given by
\begin{equation}\label{eq:integrated-mass}
    M_{tot}=\CMcal{M}\int d^3x\vert\psi\vert^2.
\end{equation}
Likewise, the mass density at any point is given by
\begin{equation}\label{eq:density}
    \rho=\frac{\CMcal{M}}{\CMcal{L}^3}\vert\psi\vert^2.
\end{equation}
All computations performed by \PyUltraLight\ are carried out using the dimensionless quantities defined in \ref{eq:dimensionless}, while provisions are made within the code for initial conditions and outputs to be converted to and from the desired unit system specified by the user. Henceforth, we will often refer to the quantity $\vert\psi\vert^2$ simply as the density, where it is understood that this is in fact a dimensionless quantity related to the physical mass density through a constant of proportionality given by equation \ref{eq:density}.

@@

Dynamical evolution within \PyUltraLight\ progresses via a symmetrised split-step Fourier process on an $N\times N\times N$ simulation grid with spatial periodic boundary conditions. To understand this method, we first consider the exact expression for the unitary time evolution of the wavefunction according to equation \ref{eq:s-adim}, namely
\begin{equation}\label{eq:exact-time-ev}
    \psi(\vec{x},t+h)=\mathcal{T}\operatorname{exp}\left[-i\int_t^{t+h}dt'\left\{-\frac{1}{2}\nabla^2+\Phi(\vec{x},t')\right\}\right]\psi(\vec{x},t),
\end{equation}
where $\mathcal{T}$ is the time-ordering symbol. For sufficiently small timestep $h$, we can use the trapezoidal rule make the approximation
\begin{equation}
    \int_t^{t+h}dt'\Phi(\vec{x},t')\approx \frac{h}{2}\Big(\Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)+\Phi(\vec{x},t)\Big).
\end{equation}
We can therefore write the approximate form of equation \ref{eq:exact-time-ev} as
\begin{equation}\label{eq:approx-time-ev}
    \psi(\vec{x},t+h)\approx \operatorname{exp}\left[i\frac{h}{2}\Big(\nabla^2-\Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)-\Phi(\vec{x},t)\Big)\right]\psi(\vec{x},t).
\end{equation}
It is acknowledged that the exponential in equation \ref{eq:approx-time-ev} omits the time-ordering symbol, and is only equivalent to its time-ordered counterpart to order $h^2$.

@@

Evaluation of equation \ref{eq:split-step} within \PyUltraLight proceeds as follows:
Initially, the nonlinear term acts in position space for one half-timestep. The result is then Fourier transformed, and a full timestep is taken in which the differential operator is applied in the Fourier domain. The potential field is then updated in accordance with equation \ref{eq:p-adim}. An inverse Fourier transform is then performed, and a final half timestep is taken in which the updated nonlinear term acts in position space to give the new $\psi$ field configuration. This procedure is known as the symmetrised split-step Fourier method, and is used widely in fields such as nonlinear fiber optics \cite{Agrawal2013}. It can be represented schematically as follows:
\begin{equation}\label{eq:schematic-psi}
    \psi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)=\operatorname{exp}\left[-\frac{ih}{2}\Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)\right]\CMcal{F}^{-1} \operatorname{exp}\left[\frac{-ih}{2}k^2\right]\CMcal{F}\operatorname{exp}\left[-\frac{ih}{2}\Phi(\vec{x},t)\right]\psi(\vec{x},t),
\end{equation}
where the order of operations is read from right to left, $\CMcal{F}$ and $\CMcal{F}^{-1}$ represent the discrete Fourier transform and its inverse, and $k$ is the wavenumber in the Fourier domain. The potential field is updated following the inverse Fourier transform in equation \ref{eq:schematic-psi}, and this update procedure may be represented schematically as
\begin{equation}\label{eq:schematic-phi}
    \Phi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)=\CMcal{F}^{-1}\left(-\frac{1}{k^2}\right)\CMcal{F}\ 4\pi\vert\psi(\vec{x},t_{i})\vert^2,
\end{equation}
where $\psi(\vec{x},t_{i})$ represents the field configuration at this intermediary point in the timestep. The final operation in equation \ref{eq:schematic-psi} changes only the phase of the $\psi$ field, and thus we could replace $\psi(\vec{x},t_{i})$ with $\psi(\vec{x},t\hspace*{-.2em}+\hspace*{-.2em}h)$ in equation \ref{eq:schematic-phi} with no change in meaning.

In the \PyUltraLight implementation, we make use of an additional simplification of the symmetrised split-step Fourier method. Namely, consecutive half-steps in the nonlinear term may be combined into a single full step in each case such that when iterated, only the first and last operations involve half steps. This can be represented schematically as 
\begin{equation}
    \psi(t\hspace*{-.2em}+\hspace*{-.2em}nh)=\operatorname{exp}\left[+\frac{ih}{2}\Phi\right]\left(\prod^n\operatorname{exp}\left[-\frac{ih}{2}\Phi\right]\operatorname{exp}\left[\frac{ih}{2}\nabla^2\right]\right)\operatorname{exp}\left[-\frac{ih}{2}\Phi\right]\psi(t),
\end{equation}
where the $\Phi$ field is updated at each step in accordance with equation \ref{eq:schematic-phi} and attention is drawn to the sign difference between the first and last operators.

The discrete Fourier transform (DFT) and its inverse are handled within \PyUltraLight using pyFFTW. pyFFTW is a pythonic wrapper around the C-based FFTW subroutine library, and allows for efficient evaluation of both real and complex DFTs \cite{pyfftw} \cite{fftw} \cite{Frigo2005}.  Multithreading is supported by pyFFTW, and is implemented within \PyUltraLight by using the multiprocessing package to determine the number of online CPUs and feed this directly to the threads argument of the pyfftw.FFTW class.

In addition to the use of pyFFTW, performance within \PyUltraLight is further improved through the use of the NumExpr package, which provides routines for the efficient manipulation of array objects within the simulation using a vector-based virtual machine \cite{numexpr}.

@@

The initialisation procedure within \PyUltraLight can be easily augmented to accommodate the requirements of the user. In the current implementation, the initial field configuration is built by loading a NumPy array file encoding a solitonic solution to the Schr{\"o}dinger-Poisson system. Using this method, solitons may be placed at the desired locations within the simulation grid, with the corresponding mass, velocity, and phase parameters each specified by the user within the accompanying Jupyter notebook.

@@

Soliton collisions can be classified into two categories depending on whether the total energy of the isolated binary system is positive or negative. In the case of positive energy, the solitons pass through each other, emerging largely undisturbed from their initial configurations following the interaction. During the collision, the wavefunctions describing the solitons are superposed, yielding distinctive interference patterns. In this section we demonstrate the ability of \PyUltraLight to reproduce the interference patterns predicted by theoretical models during binary collisions.
